### 2016-08-24 cross_compilation.tex
more corrections

This helps the \ac{OS} to recognize the location of the compiler
command everytime a new shell is opened. The \texttt{.profile} should now
contain the lines we inserted. There might be other
code in the file of other testbed administrators. We recommend to leave
other parts unmodified.
% MAC and Hostname file
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$HOME/.profile}@>
<@\textcolor{gray}{
---------------------------}
\Reactivatenumber @>
...
export TOOLCHAINDIR="$HOME/toolchains"
export TOOLCHAINBINARY="raspberry-gxx49-arm-g++"
PATH="$PATH:${TOOLCHAINDIR}/arm-rpi-4.9.3-linux-gnueabihf/bin"
...
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

This should produce a binary \texttt{hello\_world} that should only be
useful in the \ac{Raspi}. We copy it to a \ac{Raspi} to test it. We will
use \ac{SCP} to copy the binary file to one of the \ac{Raspi}s. Still, we
for example might use Fabric if we are interested in deploying a compiled
binary for many \ac{Raspi}s.

>>>>

This helps the \ac{OS} to recognize the location of the compiler
command when a new shell is opened. The \texttt{.profile} should now
contain the lines we inserted. There might be other
code in the file of other testbed administrators. We recommend to leave
other parts unmodified.
% MAC and Hostname file
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$HOME/.profile}@>
<@\textcolor{gray}{
---------------------------}
\Reactivatenumber @>
...
export TOOLCHAINDIR="$HOME/toolchains"
export TOOLCHAINBINARY="raspberry-gxx49-arm-g++"
PATH="$PATH:${TOOLCHAINDIR}/arm-rpi-4.9.3-linux-gnueabihf/bin"
...
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

This should produce a binary \texttt{hello\_world} that is executable
in the \ac{Raspi}. We copy it to a \ac{Raspi} to test it. We will
use \ac{SCP} to copy the binary file to one of the \ac{Raspi}s.
%Still, we for example might use Fabric if we are interested in deploying
Still, we might use Fabric instead if we are interested in deploying
a compiled binary for many \ac{Raspi}s.

### 2016-08-24 overlay_fs.tex
more corrections

\label{sec:overlay_fs}
In principle, our procedure modifies the image file only once in
the testbed administrator \ac{PC} when its setup is made. Also, keeping
this image in the \ac{Raspi}s provides the same initial system for all
the devices. If we do not make any further modifications during
the image setup, any files created after the initial boot of a \ac{Raspi}
will remain in the memory card. This is cumbersome to maintain since
the size of the memory card is relatively reduced (8 GB) and there
might be various users utilizing the testbed. Also, different testbed
users could be interested in running their experiments in a fresh
rebooted system with the original customized image. Nevertheless, we
emphasize that this step is not necessary if the tesbed administrator
wants to consider only persistent storage for its devices. A use case
for this scenario could be a single user for the testbed or when a testbed
administrator only wants to setup a few \ac{Raspi}s.

If both persistent and non-persistent storage are required for
the \ac{Raspi}s, here we present the steps to setup an overlay filesystem.
This type of filesystem enables an \textit{upper} filesystem to overlay
into a \textit{lower} filesystem. Whenever a file is requested, the upper
filesystem will forward the request to the lower filesystem in case it
does not have it itself. If the upper filesystem has the requested file,
it will simpy return the file. This idea can be used in our setup to mount
the root filesystem (i.e. Raspbian Lite) in the \ac{Raspi}s during startup
as read-only filesystem. Thus, on the one hand, the image configuration files
will remain after a reboot but the local data in these directories will be
erased after a hypothetic reboot. On the other hand, to enable the
possibility of persistent changes, we overlay the upper filesystem that is
mounted in the \ac{Raspi} \ac{RAM}, i.e. \texttt{/tmp} as read / writable
on top of the lower root filesystem. Thus, reading a file may return a
file from the lower filesytem, but if it is stored, it will be saved in
the upper filesystem. Accessing this file again will return
the stored file from the upper layer. After a reboot, all the stored files
in the upper filesystem will be retrieved, but the ones in the lower
filesystem that are not part of the original image will be removed.

@@

Assuming that we are still in the \texttt{chroot} environment into the
Raspbian Lite root filesystem for installing packages, we can setup the
overlay filesystem at this point of the procedure. There already exists
implementations overlaying the root filesystem. We use an implementation
available at the Git repository in \cite{overlayroot}. Since we have
installed \texttt{git} in a previous step, we clone the repository. The
command block below stores it in \texttt{/tmp} which is really mounted in
\ac{RAM}. Then, all the files stored here will disappear a \ac{Raspi} is
rebooted.

@@

Before enabling the overlaying filesystem, it is required to generate
an initial \ac{RAM} file system or \texttt{initramfs}. This is an
initial filesystem that gets loaded into \ac{RAM} during the startup
process of a Linux machine to prepare the real filesystem. For this purpose,
we need the BusyBox package by running:

@@

To generate the initial \ac{RAM} filesystem, we have to utilize the
\texttt{mkinitramfs} command. This searches by default for the available
kernel modules in the system. Since we are in \texttt{chroot} mode, we need
to specify the correct kernel modules to search for. The available
kernel modules are \texttt{/lib/modules}. To see them, we just run:

@@

Although these commands might output some warnings, they should
successfully generate working initial \ac{RAM} filesystems. Later, an
initial \ac{RAM} filesystem will need to be called by the bootloader.
In Raspbian this is done by adding a command to \texttt{config.txt} file
in the boot partition. If the system should be run in a \ac{Raspi} version 1,
then use \texttt{init.gz}, otherwise use \texttt{init-v7.gz} as follows:

%The bootloader should be made aware of the initramfs that we will generate:
\begin{lstlisting}[]
(chroot) # echo "initramfs init.gz" >> /boot/config.txt     # Only for Raspberry Pi version 1
(chroot) # echo "initramfs init-v7.gz" >> /boot/config.txt  # For Raspberry Pi version 2 or 3
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

\subsection{Persistent and Non-Persistent Image Directories}
\label{sec:non_persistent_directories}
Provided the stacked filesystem is configured, now it is possible to have
directories where files are removed or not upon rebooting the \ac{Raspi}s.
The following procedure creates an extra partition in the image for
the \ac{Raspi} user home directory that will be made storage persistent
using the upper filesystem. We first expand image according to the desired
home directory size, but avoiding to make the image bigger than the target
memory card size.

@@

We create again a loopback device and format the new partition, as follows:
\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
# mkfs.ext4 ${DEV}p3
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Again, if for some reason, the \texttt{-P} option is not available for
\texttt{losetup}, we provide an alternative command line in
Section~\ref{sec:alternative_losetup}. Finally, if the previous filesystem
formatting was successful, the filesystem is now available for use. Thus,
we need to inform the Raspbian \ac{OS} to mount the home partition
that we have just created. We will notify the \ac{OS} while
booting during the startup of the \ac{Raspi}s. This can be done by adding
an entry in \texttt{fstab} as follows:

>>>>

\label{sec:overlay_fs}
In principle, our procedure modifies the image file only once in
the testbed administrator \ac{PC} when its setup is made. Also, keeping
this image in the \ac{Raspi}s provides the same initial system for all
the devices. If we do not make any further modifications during
the image setup, any files created after the initial boot of a \ac{Raspi}
will remain in the memory card. This is cumbersome to maintain since
the size of the memory card is relatively small (8 GB) and there
might be various users utilizing the testbed. Also, different testbed
users could be interested in running their experiments in a fresh
rebooted system with the original customized image. Nevertheless, we
emphasize that this step is not necessary if the tesbed administrator
wants to consider only persistent storage for its devices. A use case
for this scenario could be a single user for the testbed or when a testbed
administrator only wants to setup a few \ac{Raspi}s.

If both persistent and non-persistent storage are required for
the \ac{Raspi}s, here we present the steps to setup an overlay filesystem.
This type of filesystem enables an \textit{upper} filesystem to overlay
into a \textit{lower} filesystem. Whenever a file is requested, the upper
filesystem will forward the request to the lower filesystem in case it
does not have it itself. If the upper filesystem has the requested file,
it will simply return the file. This idea can be used in our setup to mount
the root filesystem (i.e. Raspbian Lite) in the \ac{Raspi}s during startup
as read-only filesystem. Thus, on the one hand, the image configuration files
will remain after a reboot but the local data in these directories will be
erased after a hypothetic reboot. On the other hand, to enable the
possibility of persistent changes, we overlay the upper filesystem that is
mounted in the \ac{Raspi} \ac{RAM}, i.e. \texttt{/tmp} as read / writable
on top of the lower root filesystem. Thus, reading a file may return a
file from the lower filesytem, but if it is stored, it will be saved in
the upper filesystem. Accessing this file again will return
the stored file from the upper layer. After a reboot, all the stored files
in the upper filesystem will be retrieved, but the ones in the lower
filesystem that are not part of the original image will be removed.

@@

Assuming that we are still in the \texttt{chroot} environment into the
Raspbian Lite root filesystem for installing packages, we can setup the
overlay filesystem at this point of the procedure. There already exists
implementations overlaying the root filesystem. We use an implementation
available at the Git repository in \cite{overlayroot}. Since we have
installed \texttt{git} in a previous step, we clone the repository. The
command block below stores it in \texttt{/tmp} which is really mounted in
\ac{RAM}. Thus, all the files stored here will disappear when the system is
rebooted.

@@

Before enabling the overlaying filesystem, it is required to generate
an initial \ac{RAM} filesystem or \texttt{initramfs}. This is an
initial filesystem that is loaded into \ac{RAM} during the startup
process of a Linux machine to prepare the real filesystem. For this purpose,
we need the BusyBox package by running:

@@

To generate the initial \ac{RAM} filesystem, we have to utilize the
\texttt{mkinitramfs} command. This searches by default for the available
kernel modules in the system. Since we are in \texttt{chroot} mode, we need
to specify the correct kernel modules to search for. The available
kernel modules are located in \texttt{/lib/modules}.
To see them, we just run:

@@

Although these commands might output some warnings, they should
successfully generate working initial \ac{RAM} filesystems. Later, an
initial \ac{RAM} filesystem will need to be called by the bootloader.
In Raspbian this is done by adding a command to \texttt{config.txt} file
in the boot partition. If the system should be run in a \ac{Raspi} version 1,
then use \texttt{init.gz} by executing only the first code line below,
otherwise use \texttt{init-v7.gz} by executing only the second code line:
%as follows:

%The bootloader should be made aware of the initramfs that we will generate:
\begin{lstlisting}[]
(chroot) # echo "initramfs init.gz" >> /boot/config.txt     # For Raspberry Pi version 1
(chroot) # echo "initramfs init-v7.gz" >> /boot/config.txt  # For Raspberry Pi version 2 or 3
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

\subsection{Persistent and Non-Persistent Image Directories}
\label{sec:non_persistent_directories}
Provided the stacked filesystem is configured, now it is possible to have
directories where files are removed or not upon rebooting the \ac{Raspi}s.
The following procedure creates an extra partition in the image for
the \ac{Raspi} user home directory that will be made storage persistent.
%using the upper filesystem.
We first expand image according to the desired
home directory size, but avoiding to make the image bigger than the target
memory card size.

@@

We create a loopback device again and format the new partition, as follows:
\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
# mkfs.ext4 ${DEV}p3
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Again, if for some reason, the \texttt{-P} option is not available for
\texttt{losetup}, we provide an alternative command line in
Section~\ref{sec:alternative_losetup}. Finally, if the previous filesystem
formatting was successful, the filesystem is now available for use. Thus,
we need to inform the Raspbian \ac{OS} to mount the home partition
that we have just created.
%We will notify the \ac{OS} while booting during the startup of the
%\ac{Raspi}s.
This can be done by adding
an entry in \texttt{fstab} as follows:

### 2016-08-24 tools.tex
more corrections

The previous \texttt{fabfile} shows three function that perform our
example tasks. These functions utilize variables and subsequent functions
from the Fabric \ac{API} such as \texttt{env}, \texttt{task} and \texttt{sudo}
among others. Each of these \ac{API} functions permit to define environment
variables, create the administrator tasks through decorators or run the
mentioned task in \texttt{sudo} mode, respectively.
When a task is called from the terminal, Fabric searches the directory for
the \texttt{fabfile.py} file and executes the desired task. The syntax for
executing a task with arguments is in the form
\texttt{fab <TASK>:arg1,arg2,...}. In what follows, we denote the
\ac{IP} address of a generic \ac{Raspi} for test as \texttt{<RASP\_IP>}.
The executions from the terminal of some of these commands are shown as
follows:

@@

There are methods to enable the \ac{Raspi}s to continue running
applications although the connection is terminated either on purpose or
unexpectedly. One method is to run programs within a \texttt{screen} session.
\texttt{screen} enables a user to run applications within a shell
window, a \texttt{screen} session, which does not terminate even with
connectivity interruptions. Users can attach and detach from such a session
as desired. The following procedure presents how to use \texttt{screen} with
\ac{SSH} to: (i) Login to a generic \ac{Raspi}, (ii) open a screen session,
(iii) execute an example command, (iv) detach from the screen session, (v)
terminate the \ac{SSH} connection, (vi) login to the \ac{Raspi} again
and (vii) attach to screen session to see the program still running.
From the testbed administrator, we start by establishing \ac{SSH} connection
to a \ac{Raspi} and open a screen session:

@@

From the command output, we see that the session is still running in our
generic \ac{Raspi} number \texttt{XX} and that no user is currently attached
to the session. To attach to the session, we execute:

@@

After attaching again, we should see the \texttt{top} still running.
\texttt{screen} has more functionalities that can be used in this
or other contexts but that is outside the scope of this work.
To terminate the \texttt{screen} session, first terminate \texttt{top}
by pressing \texttt{q} in the keyboard. Once out of the \texttt{top}
output, we need to type \texttt{exit} two times to first exit the
\texttt{screen} session and then terminate the \ac{SSH} connection.
An output should be as follows:

>>>>

The previous \texttt{fabfile} shows three functions that perform our
example tasks. These functions utilize variables and subsequent functions
from the Fabric \ac{API} such as \texttt{env}, \texttt{task} and \texttt{sudo}
among others. Each of these \ac{API} functions permit to define environment
variables, create the administrator tasks through decorators or run the
mentioned task in \texttt{sudo} mode, respectively.
When a task is called from the terminal, Fabric searches the directory for
the \texttt{fabfile.py} file and executes the desired task. The syntax for
executing a task with arguments is in the form
\texttt{fab <TASK>:arg1,arg2,...}. In what follows, we denote the
\ac{IP} address of a generic \ac{Raspi} for test as \texttt{<RASP\_IP>}.
The executions from the terminal of some of these commands are shown as
follows:

@@

There are methods to enable the \ac{Raspi}s to continue running
applications although the connection is terminated either on purpose or
unexpectedly. One method is to run programs within a \texttt{screen} session.
\texttt{screen} enables a user to run applications within a shell
window, a \texttt{screen} session, which does not terminate even with
connectivity interruptions. Users can attach and detach from a
\texttt{screen} session
as desired. The following procedure presents how to use \texttt{screen} with
\ac{SSH} to: (i) Login to a generic \ac{Raspi}, (ii) open a screen session,
(iii) execute an example command, (iv) detach from the screen session, (v)
terminate the \ac{SSH} connection, (vi) login to the \ac{Raspi} again
and (vii) attach to screen session to see the program still running.
From the testbed administrator \ac{PC}, we start by establishing a \ac{SSH}
connection to a \ac{Raspi} and open a screen session:

@@

From the command output, we can see that the session is still running in our
generic \ac{Raspi} number \texttt{XX} and that no user is currently attached
to the session. To attach to the session, we execute:

@@

After attaching again, we should see \texttt{top} still running.
\texttt{screen} has more functionalities that can be used in this
or other contexts but that is outside the scope of this work.
To terminate the \texttt{screen} session, first terminate \texttt{top}
by pressing \texttt{q} in the keyboard.
Once \texttt{top} is terminated,
%Once out of the \texttt{top} output,
we need to type \texttt{exit} two times to first exit the
\texttt{screen} session and then terminate the \ac{SSH} connection.
An output should be as follows:
